---
title: "HMM with INLA"
author: "G. Nuel"
date: '`r format(Sys.time(), "%d %B, %Y")`'
#header-includes:
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: false
  word_document: default
#bibliography: biblio_AD.bib
#biblio-style: apalike
editor_options: 
  chunk_output_type: console
---

# bandlinear systems

The package $\texttt{bandsolve}$ to solve efficiently $Ax=b$ where $A$ is a symmetric band matrix. Complexity in $\mathcal{O}(n L)$ where $L$ is the number of bands. Also usable to obtain the inverse of $A$ (by solving with $b$ base vectors).

```{r message=FALSE}
#install.packages("remotes")
#remotes::install_github("Monneret/bandsolve")
require(bandsolve)
n=20
A=matrix(0,n,2)
A[,1]=3.0; A[-n,2]=-1.0
b=rep(1.0,n)
x=bandsolve(A,b)
norm(rot2mat(A)%*%x-b)

#bandsolve_cpp(A,as.matrix(b))

```

When there is only one band off diagonal, the determinant of $A$ can also be computed efficiently using the recursion:
$$
D_i=A[i,1]\times  D_{i+1}-A[i,2]^2 \times D_{i+2}
$$
where $D_i=\det(A[i:n,i:n])$. Log-scale implementation possible.

```{r}

# log determinant
A=matrix(0,n,3)
A[,1]=3.0; A[-n,2]=-1.0; A[1:(n-2),3]=+1.0

bandlogdet=function(A) {
  # copy matrix to avoid in-place computation
  Amem=matrix(NA,nrow(A),ncol(A))
  Amem[]=A[]
  # call LDL
  invisible(LDL(Amem))
  # return res
  return(sum(log(Amem[,1])))
}

c(log(det(rot2mat(A))),bandlogdet(A))
```



# Gaussian order 1 random walk

## Model

$$
\mathbb{P}(X_1=x_1)\propto 1
\quad
\mathbb{P}(X_i=x_i|X_{i-1}=x_{i-1})=\texttt{dnorm}(x_i,x_{i-1},\sigma)
\quad
\mathbb{P}(Y_i=y_i | X_i=x_i)=\texttt{dnorm}(y_i,x_{i},\varepsilon)
$$

## Derivatives

$$
f(x|y,\theta)=\log \texttt{dnorm}(x_1) + \sum_{i=2}^n \log \texttt{dnorm}(x_i,x_{i-1},\sigma) + \sum_{i=1}^n \texttt{dnorm}(y_i,x_{i},\varepsilon)
$$
$$
\mathrm{grad}f(x | y,\theta)=\left(
-x_1 + \frac{(x_2-x_1)}{\sigma^2} + \frac{(y_1-x_1)}{\varepsilon^2} ;
\frac{(x_i-x_{i-1})}{\sigma^2} + \mathbf{1}_{i<n} \frac{(x_{i+1}-x_{i})}{\sigma^2} + \frac{(y_i-x_i)}{\varepsilon^2}
\quad i=2,\ldots,n
\right)^T
$$
$$
\mathrm{Hess}f(x|y,\theta)=\mathrm{band}(A)
$$
$$
A[,1]=\left(
-1-\frac{1}{\sigma^2}-\frac{1}{\varepsilon^2} ;
-\frac{1}{\sigma^2}-\mathbf{1}_{i<n}\frac{1}{\sigma^2}-\frac{1}{\varepsilon^2}\quad i=2,\ldots,n
\right)^T
\quad
A[,2]=\left(
\frac{1}{\sigma^2} \quad i=1,\ldots,n-1 ;
0
\right)^T
$$

## Newton-Raphson for MAP

$$
\widehat{x}^\theta = \arg \max_x f(x | y,\theta)
$$
$$
x^{(0)}=0 \quad \text{iter}=1\ldots,N: x^{(\text{iter})}=x^{(\text{iter}-1)} - \mathrm{Hess}f\left(x^{(\text{iter}-1)}|y,\theta\right)^{-1} \mathrm{grad}f\left(x^{(\text{iter}-1)}|y,\theta\right)
$$

## Laplace approximation

$$
\log \mathbb{P}(y | \theta) \simeq \text{cst.} + \log \mathbb{P}\left(
\widehat{x}^\theta,y|\theta \right)
- \frac{1}{2} \log \det \left(
-\mathrm{Hess}f\left(\widehat{x}^{\theta}|y,\theta\right)
\right)
$$

## Gaussian posterior of $x$

$$
\mathbb{P}\left(x | y,\theta\right) = \texttt{dmvnorm}\left(
\widehat{x}^{\theta},\left(
-\mathrm{Hess}f\left(\widehat{x}^{\theta}|y,\theta\right)
\right)^{-1}
\right)
$$

## Implementation

```{r eval=TRUE,echo=FALSE,fig.cap="simulation of a order 1 Gaussian random walk"}
n=500
eps=2.0
sigma=0.5
theta_star=log(c(eps,sigma))
theta=theta_star

set.seed(42)

x=y=rep(NA,n)
#x[1]=rnorm(1)
x[1]=7.0
for (i in 2:n)
  x[i]=rnorm(1,x[i-1],sigma)
y=rnorm(n,x,eps)

x_star=x

plot(x_star,t='l',ylim=range(y),xlab="i",ylab="",lwd=2)
points(y,cex=0.2)
legend("topright",c("hidden state","observations"),pch=c(NA,1),lty=c(1,NA),lwd=c(2,1),bg="white")

#plot(y,cex=0.2)
```


```{r echo=FALSE,fig.cap="posterior hidden path with MLE parameters"}
f=function(x,theta) {
  eps=exp(theta[1])
  sigma=exp(theta[2])
  #dnorm(x[1],log=TRUE)+sum(dnorm(x[-1],x[-n],sigma,log=TRUE))+sum(dnorm(y,x,eps,log=TRUE))
  sum(dnorm(x[-1],x[-n],sigma,log=TRUE))+sum(dnorm(y,x,eps,log=TRUE))
}

grad=function(x,theta) {
  eps=exp(theta[1])
  sigma=exp(theta[2])
  # grad
  b=(y-x)/eps^2
  i=1:n
  b[i[-1]]=b[i[-1]]-diff(x)/sigma^2
  b[i[-n]]=b[i[-n]]+diff(x)/sigma^2
  return(b)
}
# verif
# i=42
# h=rep(0,n); h[i]=1e-5;
# c(
#   (f(x+h,theta)-f(x,theta))/1e-5,
#    grad(x,theta)[i]
# )

Hessian=function(theta) {
  eps=exp(theta[1])
  sigma=exp(theta[2])
  # Hessian
  A=matrix(0.0,n,2)
  i=1:n
  A[,1]=-1/eps^2
  A[i[-1],1]=A[i[-1],1]-1/sigma^2
  A[i[-n],1]=A[i[-n],1]-1/sigma^2
  A[-n,2]=1/sigma^2
  return(A)
}
# verif
# i=1
# h=rep(0,n); h[i]=1e-5;
# aux=(grad(x+h,theta)-grad(x,theta))/1e-5
# aux[aux!=0]
# Hessian(theta)[i,]


xhat=function(theta,itermax=50,tol=1e-5) {
  # parameters
  # Newton-Raphson loop
  x=rep(1.0,n)
  for (iter in 1:itermax) {
    delta=bandsolve(Hessian(theta),grad(x,theta))
    x=x-delta
    if (max(abs(delta)/abs(x))<tol) break
  }
  return(x)
}

loglik=function(theta) {
  x=xhat(theta)
  A=Hessian(theta)
  f(x,theta)-1/2*bandlogdet(-A)
}

# finding MLE
opt=optim(par=theta_star,fn=function(theta){-loglik(theta)})
rbind(opt$par,theta_star)

x=xhat(opt$par)
sd=sqrt(diag(bandsolve(-Hessian(opt$par))))

plot(x_star,t='l',ylim=range(y),xlab="i",ylab="",lwd=2)
points(y,cex=0.2)
legend("topright",c("hidden state","observations","estimation"),pch=c(NA,1,NA),lty=c(1,NA,1),lwd=c(2,1,2),col=c("black","black","blue"),bg="white")
points(x,t='l',col="blue",lwd=2)
polygon(c(1:n,n:1),c(x-1.96*sd,rev(x+1.96*sd)),col=rgb(0,0,1,alpha=0.30),border=FALSE)
```

# derivative of the MAP

The objective is to compute the gradient of $\widehat{x}^\theta$ with respect with $\theta$.The basic idea is that whe can obtain a good approximation of $\widehat{x}^{\theta+h}$ from $\widehat{x}^\theta$ with just one Newton-Raphson step:

$$
\widehat{x}^{\theta+h} \simeq \hat{x}^\theta
- \mathrm{Hess}f\left(\widehat{x}^\theta|y,\theta+h\right)^{-1}
\mathrm{grad}f\left(\widehat{x}^\theta|y,\theta+h\right)
$$

The second idea is to assume that the Hessian can be computed in $\theta$ instead of $\theta+h$ since it is not changing a lot if $h$ is small. We denote with $\nabla$ the gradient with respect with $\theta$ and we obtain the following approximation:

$$
\nabla\widehat{x}^{\theta} \simeq
- \mathrm{Hess}f\left(\widehat{x}^\theta|y,\theta\right)^{-1}
\nabla \mathrm{grad}f\left(\widehat{x}^\theta|y,\theta\right)
$$

NB: we have a closed form formulat for the $\nabla$ of the gradient so it is easy.

```{r}
theta=opt$par

h=1e-5
thetah1=theta+c(h,0)
thetah2=theta+c(0,h)
x1=xhat(thetah1)
x2=xhat(thetah2)

ggrad=function(x,theta) {
  eps=exp(theta[1])
  sigma=exp(theta[2])
  # grad
  b1=(y-x)/eps^2
  b2=rep(0,n)
  i=1:n
  b2[i[-1]]=b2[i[-1]]-diff(x)/sigma^2
  b2[i[-n]]=b2[i[-n]]+diff(x)/sigma^2
  return(-2*cbind(b1,b2))
}

ref=cbind((x1-x)/h,(x2-x)/h)
res=-bandsolve(Hessian(theta),ggrad(x,theta))

plot(c(ref),c(res),cex=0.5); abline(0,1,col="blue")
```

