---
title: "Bandsolve for INLA"
author: "G. Nuel"
date: '`r format(Sys.time(), "%d %B, %Y")`'
#header-includes:
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: false
  word_document: default
#bibliography: biblio_AD.bib
#biblio-style: apalike
editor_options: 
  chunk_output_type: console
---

The package $\texttt{bandsolve}$ to solve efficiently $Ax=b$ where $A$ is a symmetric band matrix. Complexity in $\mathcal{O}(n L)$ where $L$ is the number of bands. Also usable to obtain the inverse of $A$ (by solving with $b$ base vectors).

```{r message=FALSE}
#install.packages("remotes")
#remotes::install_github("Monneret/bandsolve")
require(bandsolve)
n=20
A=matrix(0,n,2)
A[,1]=3.0; A[-n,2]=-1.0
b=rep(1.0,n)
x=bandsolve(A,b)
norm(rot2mat(A)%*%x-b)

#bandsolve_cpp(A,as.matrix(b))

```

When there is only one band off diagonal, the determinant of $A$ can also be computed efficiently using the recursion:
$$
D_i=A[i,1]\times  D_{i+1}-A[i,2]^2 \times D_{i+2}
$$
where $D_i=\det(A[i:n,i:n])$. Log-scale implementation possible.

The log of the determinant is also easy to obtain:

```{r}

# log determinant
A=matrix(0,n,3)
A[,1]=3.0; A[-n,2]=-1.0; A[1:(n-2),3]=+1.0

bandlogdet=function(A) {
  # copy matrix to avoid in-place computation
  Amem=matrix(NA,nrow(A),ncol(A))
  Amem[]=A[]
  # call LDL
  invisible(LDL(Amem))
  # return res
  return(sum(log(Amem[,1])))
}

c(log(det(rot2mat(A))),bandlogdet(A))
```



